\section{Kleisli Categories}

Managing side-effects, and separating concerns.

\subsection{Writer Category}

\begin{example}
    To abstract concerns such as logging, it may be possible to utilize \textit{embellished} or \textit{decorated} functions.
    
    For instance, given the embellished function \mintinline{typescript}{isEven} which not only carries out the primary predicate functionality, it also produces a log message as output. Together, they are returned as a 2-tuple
    \begin{minted}{typescript}
function isEven(n: number): [bool, string] {
    return [ n % 2 === 0, "isEven(number) called" ];
}
    \end{minted}
    
    Given another embellished function \mintinline{typescript}{negate} which also produces a log message
    \begin{minted}{typescript}
function negate(b: bool): [ bool, string ] {
    return [ !b, "Negate(bool) called" ];
}
    \end{minted}
    
    Their composition would not be via \mintinline{hs}{negate . isEven} since their input output types mismatch; but, something like
    \begin{minted}{typescript}
function isOdd(n: number): [ bool, string ] {
    const [ pred, log1 ] = isEven(n);
    const [ neg, log2 ] = negate(pred);
    return [ neg, log1 + log2 ];
}
    \end{minted}
\end{example}

\begin{definition}[Composing Embellished Functions]
    The steps to composing such embellished functions are:
    \begin{enumerate}
        \item Execute embellished function of first morphism.
        \item Extract first component of result pair, pass it to embellished function of second morphism.
        \item Concatenate second component of first result and second component of second result.
        \item Construct and return new pair with first component of final result with concatenated second components.
    \end{enumerate}
\end{definition}

\begin{remark}
    In TypeScript, this would be implemented via something like
    \inputminted{typescript}{content/code-listings/writer-compose.ts}
    
    Then the two embellished functions \mintinline{typescript}{negate} and \mintinline{typescript}{isEven} can now be composed
    
    \begin{minted}{typescript}
const isOdd: number => Writer<bool> = 
    composeW(isEven, negate);
    \end{minted}
\end{remark}

\begin{remark}
    In the case of \mintinline{typescript}{Writer}s, needing to compose them creates the need for identity morphisms.
\end{remark}

\begin{definition}[Identity Embellished Function]
    The embellished identity function takes the form
    \inputminted{typescript}{content/code-listings/writer-id.ts}
\end{definition}
